The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-10 paper/ms.tex
Small changes throughout

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the computed values of $J_z$ for the thick-disk (top
panel) and thin-disk (bottom panel) orbits assuming a $z$ offset of
$100\,\pc$, at different orbital phases. One can see that if the observed $z$ values have a bias (from e.g.
an incorrectly computed midplane), then the induced error distribution in
$J_z$ is decidedly non-Gaussian. Therefore, any sort of error propagation must
take this into account. Intuition for the shape of each panel is given in the
text. We also plot one half the 95\uth{} percentile minus the 5\uth{}
percentile of each distribution as a horizontal arrow anchored on the true
$J_z$ value. We call this $\Delta J_z$ and will use it (along with the
similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to empirically describe
the error distribution. We see that $\Delta J_z$ roughly corresponds to the
distance from the true $J_z$ value to one of the modes of the distribution of
computed $J_z$ values. Similar plots for $J_R$ induced by a $z$ offset and
$J_R$ and $J_{\phi}$ induced by an $x$ offset are given in
Appendix~\ref{app:hist}.}
\label{fig:Jz_hist}
\end{figure}

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the values of
$J_z$ computed for the thick-disk orbit (top panel) and thin-disk orbit (bottom panel)
assuming a $z$ offset of $100\,\pc$ (i.e. as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}), as a function of orbital phase. The true value of $J_z$ is plotted as a
vertical dashed line. The systematic error in $J_z$ induced by a
systematic offset in $z$ is non-Gaussian and bimodal; neither of the modes is centered on the true $J_z$ (i.e. the $J_z$ computed with no offset in $z$). In the case of the thin-disk orbit
(bottom panel), we see that, in addition to the prior complications, the
distribution is not even centered on the true value. This comes about when the midplane
error is $\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit, equivalent to $A_z$ in the epicyclic approximation (see
Section~\ref{ssec:epi_action}).

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(-190, 0, v_z)\,\kms$, where we vary $v_z$. We show
this for a $z$ offset of $10, 50, 100\,\pc$ (orange, teal, and green
respectively). As before, the error ($\Delta J_z$) is one half the 95\uth{}
minus 5\uth{} percentile of the distribution of $J_z$ values over the
course of the orbit. There are large errors for thin-disk like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(-190, 0, v_z)\,\kms$, where
we vary $v_z$. For a thin disk-like orbit ($J_z\sim0.5\,\actunit$), even a
$10\,\pc$ offset in $z$ is enough to induce a $\sim25\%$ error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. (\Gus{Check:}) The upper panels
show the midplane as a function of azimuth before the first encounter near the
Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$. As later encounters occur
which bring the Sagittarius-like object closer to the Solar circle, we can see
that stronger midplane variations result. In fact, one can see that an $m=2$
mode is excited in the lower left panel at $t=6.0\,\Gyr$ --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited (note that $m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction). The fact that the $t=6.9\,\Gyr$
panel looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) suggests that midplane variations are
produced at least in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

\subsubsection{Galactic Center position}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to \sgra{}, referred to as $R_0$, can be
precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}, with a
recent measurement using near-infrared interferometry of $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}, or $\sim0.3\%$. However, the location of \sgra{} may not be equivalent to the location of the ``dynamical Galactic center,'' i.e., the
point in three-dimensional space about which the stars in the solar
neighborhood are orbiting. This assumption, although sensible and frequently
made, has not yet been justified.

@@

We did not consider in this text the effect of the \emph{angular position} of
the dynamical Galactic center being offset with \sgra{}.

@@

\vspace{6pt}

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the MW with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim$185, 162, 84 pc (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim$30 and $\sim$100 pc before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the MW suggests that these values are consistent with expectations for midplane variations in the MW itself, which have not yet been measured as a function of azimuth.

>>>>

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_Jz_zerr_hist.pdf}
\end{center}
\caption{A histogram of the computed values of $J_z$ at different orbital phases for the thick-disk (top
panel) and thin-disk (bottom panel) orbits assuming a $z$ offset of
$100\,\pc$. One can see that if the observed $z$ values have a bias (from e.g.
an incorrectly computed midplane), then the induced error distribution in
$J_z$ is decidedly non-Gaussian. Therefore, any sort of error propagation must
take this into account. Intuition for the shape of each panel is given in the
text. We also plot one half the 95\uth{} percentile minus the 5\uth{}
percentile of each distribution as a horizontal arrow anchored on the true
$J_z$ value. We call this $\Delta J_z$ and will use it (along with the
similarly defined $\Delta J_R$ and $\Delta J_{\phi}$) to empirically describe
the error distribution. We see that $\Delta J_z$ roughly corresponds to the
distance from the true $J_z$ value to one of the modes of the distribution of
computed $J_z$ values. Similar plots for $J_R$ induced by a $z$ offset and
$J_R$ and $J_{\phi}$ induced by an $x$ offset are given in
Appendix~\ref{app:hist}.}
\label{fig:Jz_hist}
\end{figure}

@@

In Figure~\ref{fig:Jz_hist} we plot a histogram of the values of
$J_z$ computed at different orbital phases for the thick-disk orbit (top panel) and thin-disk orbit (bottom panel)
assuming a $z$ offset of $100\,\pc$ (as in the upper right panel of
Figure~\ref{fig:one_orbit_wrong_ref}). The true value of $J_z$ is plotted as a
vertical dashed line. The systematic error in $J_z$ induced by a
systematic offset in $z$ is non-Gaussian and bimodal; neither of the modes is centered on the true $J_z$ (i.e. the $J_z$ computed with no offset in $z$). In the case of the thin-disk orbit
(bottom panel), we see that, in addition to the prior complications, the
distribution is not even centered on the true value. This comes about when the midplane
error is $\gtrsim z_{\text{max}}$, where $z_{\text{max}}$ is the maximum
height of the orbit, equivalent to $A_z$ in the epicyclic approximation (see
Section~\ref{ssec:epi_action}).

@@

\begin{figure}
\begin{center}
\includegraphics[width=\columnwidth]{fig/schmactions_many_orbits_Jz_fun.pdf}
\end{center}
\caption{The fractional error in $J_z$ as a function of $J_z$ for a few
different offsets in $z$. All orbits have the same initial position of $(8, 0,
0)\,\kpc$ and velocity $(0, -190, v_z)\,\kms$, where we vary $v_z$. We show
this for a $z$ offset of $10, 50, 100\,\pc$ (orange, teal, and green
respectively). As before, the error ($\Delta J_z$) is one half the 95\uth{}
minus 5\uth{} percentile of the distribution of $J_z$ values over the
course of the orbit. There are large errors for thin-disk like orbits ($J_z
\sim 0.5\,\actunit$), even for a small midplane offset of $10\,\pc$. As dashed
lines in each color we also plot the prediction for $\Delta J_z/J_z$ from the
epicyclic approximation (Equation~\eqref{eq:Ji_err_mosttime}), which shows
excellent agreement with the numerically computed values.}
\label{fig:dJz_fun_Jz}
\end{figure}

To further understand the effect of the midplane error, we also plot the
fractional error in $J_z$ as a function of $J_z$ for $z$ offsets of $10, 50,
100\,\pc$ (orange, teal, and green, respectively) in
Figure~\ref{fig:dJz_fun_Jz}. For each orbit, we set the initial position to be
$(8,0,0)\,\kpc$ and the initial velocity to be $(0, -190, v_z)\,\kms$, where
we vary $v_z$. For a thin disk-like orbit ($J_z\sim0.5\,\actunit$), even a
$10\,\pc$ offset in $z$ is enough to induce a $\sim25\%$ error in $J_z$. For
larger values of $J_z$, the fractional errors are suppressed, but the induced
error can still be large depending on how great the $z$ offset is. We also
plot the epicylic prediction for $\Delta J_z / J_z$ from
Equation~\eqref{eq:Ji_err_mosttime} as dashed lines for each $z$ offset. We
find that the epicyclic approximation matches the numerical estimate quite
well.

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=342.078286667pt]{fig/midplane_fit_chervinsim.pdf}
\end{center}
\caption{The local midplane determined at the fiducial Solar radius
($8.2\,\kpc$) for 4 different time steps from live N-body simulations of a
Sagittarius encounter with the Milky Way \citep{2018MNRAS.481..286L}. As
before, we have subtracted the best fit $A\sin{(\phi+B)}+C$ curve to account
for inaccuracies in the galactocentric coordinate system. Error bars are
calculated as in Figure~\ref{fig:midplane}. The upper panels
show the midplane as a function of azimuth before the first encounter near the
Solar circle at $t=2.0\,\Gyr$ and $t=4.0\,\Gyr$. As later encounters occur
which bring the Sagittarius-like object closer to the Solar circle, we can see
that stronger midplane variations result. In fact, one can see that an $m=2$
mode is excited in the lower left panel at $t=6.0\,\Gyr$ --- this is
consistent with Figure~17 of \citet{2018MNRAS.481..286L}, which shows that an
$m=2$ mode is excited ($m=0$ and $m=1$ modes are stronger, but these
are removed in our sine-curve subtraction). The fact that the $t=6.9\,\Gyr$
panel looks qualitatively similar to the panels from the FIRE simulations
(Figure~\ref{fig:midplane}) suggests that midplane variations are
produced at least in part by mergers.}
\label{fig:midplane_chervin}
\end{figure*}

@@

\subsubsection{Galactic Center Position}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to \sgra{}, referred to as $R_0$, can be
precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}, with a
recent measurement using near-infrared interferometry of $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}, or $\sim0.3\%$. However, the location of \sgra{} may not be equivalent to the location of the ``dynamical Galactic center,'' i.e., the
point in three-dimensional space about which the stars in the solar
neighborhood are orbiting. This assumption, although sensible and frequently
made, has not yet been justified.

@@

We did not consider in this work the effect of the \emph{angular position} of
the dynamical Galactic center being offset with \sgra{}.

@@

@@

\item To demonstrate the previous point, we measured the local galactic
midplane along the Solar circle in three different high-resolution, zoom-in
simulations of Milky Way mass galaxies from the FIRE collaboration, as well as a controlled simulation of the interaction of the Milky Way with Sagittarius. We found
that the midplane varies as a function of azimuth at the Solar circle by
$\sim$185, 162, 84 pc (middle $90\%$) for the three cosmological simulations we considered
(\mi{}, \mf{}, and \mm{}, respectively) and by $\sim$30 and $\sim$100 pc before and after the interaction in the controlled simulation. Comparison of the corresponding velocity variations with recent measurements in the Milky Way suggests that these values are consistent with expectations for midplane variations in the Milky Way itself, which have not yet been measured as a function of azimuth.
